Arterial pulsations cannot drive intramural periarterial drainage: Significance for Alzheimer's disease

### 2017-07-18 paper.tex
frontiers submission

\documentclass[a4paper,titlepage]{scrartcl}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage[format=hang,margin=1cm]{caption}
\usepackage{rotating}
\usepackage{subcaption}
\usepackage{authblk}
\usepackage{color}
\usepackage{xspace}
\usepackage{textgreek}

\begin{document}

\input{definitions}

\sisetup{range-phrase=--}
\sisetup{range-units=single}

% ----------------------------------------------------------------------------
% Titlepage Parameters
% ----------------------------------------------------------------------------

\title{Arterial pulsations are not powerful enough to drive perivascular drainage of solutes from the brain}
\author[1,2]{Alexandra K. Diem}
\author[2]{Neil W. Bressloff}
\author[3]{Roxana O. Carare}
\author[4]{Giles Richardson}
\affil[1]{Institute for Complex Systems Simulation, University of Southampton, Southampton SO17 1BJ, UK}
\affil[2]{Computational Engineering and Design, Faculty of Engineering \& the Environment, University of Southampton, Southampton SO17 1BJ, UK}
\affil[3]{Faculty of Medicine, University of Southampton, Southampton, UK}
\affil[4]{Faculty of Mathematical Sciences, University of Southampton, Southampton, UK}

@@

\begin{abstract}
\textbf{Summary.} Alzheimer's Disease (AD) is the most common form of dementia and to date there is no preventive medication. Its development is closely related to the accumulation of the neuronal metabolite \Ab, which raises the question of how metabolites are eliminated from the brain. Unlike the rest of the body the brain does not have a lymphatic system and therefore has to eliminate metabolites differently. A connection between the pathology of AD and cerebrovascular diseases has been suggested and confirmed in various mice studies, which has lead to the assumption that interstitial fluid in the basement membranes of cerebral arteries provide the pathways for the perivascular drainage of \Ab. However, perivascular drainage occurs in the reverse direction to arterial blood flow, an effect that lacks an effective explanation. In this study we tested the well-accepted hypothesis that arterial pulsations drive perivascular drainage and additionally that a valve mechanism exists which ensure net reverse drainage. This was examined in terms of a  mathematical model of the  drainage mechanism. It is shown that a valve mechanism (such as directional permeability of the perivascular space) is necessary to achieve a net reverse flow. However, it is also shown that arterial pulsations are not strong enough to produce drainage velocities comparable to experimental observations. % 209 words

Word count: 209

\vspace{1cm}

\textbf{Keywords:} Alzheimer's disease, brain lymphatics, cerebrovascular flow
\end{abstract}

% ----------------------------------------------------------------------------
% Paper Body
% ----------------------------------------------------------------------------

% Introduction

Dementia, a term often used synonymously with Alzheimer's disease (AD), is a symptom of a variety of diseases affecting the human brain and does not constitute a part of normal ageing. Although it has been studied for over 100 years \cite{Selkoe2001}, to date there is no cure as the processes relevant to its onset and progression have not yet been fully understood. The pathology of AD comprises different characteristics, some of which are degeneration and death of neurons, synapse loss, neuroinflammation and the accumulation of waste product deposits \cite{Koffie2011}. The protein amyloid-$\beta$ (\Ab) is produced as part of the normal metabolism of healthy brains, but it plays a major role in the development of AD \cite{Haass1992}. Its accumulation in the walls of cerebral blood vessels as cerebral amyloid angiopathy (CAA) is closely associated with the onset and progression of AD \cite{Weller2006, Weller2009}. CAA strongly correlates with the progression of AD and the degree of dementia as it appears in \SIrange{90}{96}{\percent} of AD patients \cite{Weller2009a}.

In order to process metabolites, the brain requires a lymphatic system that transports waste products to the lymph nodes, where they are decomposed. While the rest of the body possesses a dedicated network of lymphatic vessels these do not exist in the brain \cite{Weller2010}. There is ample evidence that the basement membranes (BM) in the walls of cerebral arteries provide the clearance pathway for interstitial fluid (ISF) towards the surface of the brain, also referred to as perivascular drainage \cite{Weller2010,Carare2008,Hawkes2011,Morris2014}. Because it provides the pathways for perivascular drainage, the BM is also referred to as perivascular space (PVS) in this context. It is important not to confuse this PVS with the Virchow-Robin spaces, which are also often referred to as perivascular (or ``paravascular'') spaces. However, Virchow-Robin spaces are virtual spaces and have no role in \Ab drainage. Additionally, the BM does not actually constitute a space, but rather a fluid-filled protein matrix. Our understanding of the mechanism that drives perivascular drainage through the BM is still very limited. The evidence indicates that ISF enters the BM at the capillary level and flows towards the lymph nodes in the neck via cerebral arteries, but not veins. This suggests that ISF flow is driven in the reverse direction of the blood flow, a phenomenon that does not seem very plausible at first sight \cite{Carare2008}.

A widely accepted hypothesis for the driving mechanism of this reverse perivascular drainage of solutes from the brain is that it is driven by arterial pulsations \cite{Weller2009,Carare2008,Hawkes2011,Morris2014,Schley2006,Attems2011,Wang2011,Iliff2012,Asgari2015,Sharp2015,Weller2015a}. In this paper we tested this hypothesis by using a multi-scale modelling approach combining analytical and numerical methods. The BM is modelled as a porous medium using Darcy's law and a lubrication approximation. Flow inside the BM is driven by both the ISF pressure gradient and the deformation of the arterial wall, which are obtained from a numerical model of the middle cerebral artery (MCA). To guarantee reverse drainage (\ie in the opposite direction of the blood flow) a valve mechanism is implemented as a pressure gradient dependent permeablility. It is shown that this valve mechanism is neccesary to achieve net reverse drainage. The results show that this valve mechanism successfully ensures net reverse drainage inside the BM and is neccesary. However, due to the long wavelength of arterial pulsations the pressure gradients inside the BM were found to be too small to drive fluid flow at the expected velocities. The resulting velocity of the model was approximately two orders of magnitude smaller than values obtained experimentally \cite{Carare2008}. Therefore arterial pulsations are concluded to be insufficient to drive perivascular drainage along the BM of cerebral arteries.

\section{Porous medium model of lymphatic drainage in the BM}
\label{sec:diodemodel}

We propose a model of lymphatic drainage from the brain via the cerebral BM based on Darcy's law for flow in porous media. By exploiting the discrepancy of scales between the width of the BM and its length we approximate the flow by a one-dimensional lubrication model (see for example \cite{Ockendon1995}). Furthermore, we allow the permeability of the BM to depend on the pressure gradient of the ISF. Such a pressure gradient dependent permeability could, for instance, model proteins in the BM protein network that have structural properties that allow them to bend easily only in one direction, thereby acting as a valve, and thus provide a higher resistance to flow in the direction of the blood flow. A similar idea has been investigated in \cite{Sharp2015}, where proteins were explicitly modelled as cillia. However, here we model the active component of the BM generically in this model and thus allow for alternative explanations of the phenomenon.

The BM consists of a complex matrix of proteins (extracellular matrix, ECM) that can be interpreted as a porous medium through which soluble metabolites drain via the ISF. We model flow through this medium, in the standard fashion, by Darcy's law in a 2D cylindrical coordinate system. Darcy's law states that the flux $\q$ is related to the fluid pressure gradient $\boldsymbol{\nabla} p$ by
\beq\label{eq:darcy}
  \q = -\frac{k}{\mu} \boldsymbol{\nabla} p
\eeq
where $k$ is the intrinsic permeability of the ECM and $\mu$ is the viscosity of ISF. Since ISF flows up the artery, counter to the direction of arterial pulsations, a valve-like mechanism in the ECM is required if arterial pulsations are indeed the motive force for this flow. This valve-like property of the ECM can be modelled by a Darcy flow \eqref{eq:darcy}, in which the intrinsic permeability is a function of the axial pressure gradient $\dd p / \dd z$ along the vessel. On assuming an artery aligned along the $z$-axis with net blood flow and arterial pulsations propagating the positive $z$-direction we model valve-like properties of the ECM in the BM by a pressure gradient dependent permeability
\beq
k = \mu K(p_z) \qquad \text{where } p_z = \frac{\dd p}{\dd z}.
\eeq
In order to obtain a net flux up the vessel, counter to the blood flow, we require this function to be increasing in $p_z$, representing the notion that it is harder to push a flow in the positive $z$-direction ($p_z < 0$ and $K(p_z)$ small) than in the negative $z$-direction ($p_z \geq 0$ and $K(p_z)$ large).

The governing equations in cylindrical coordinates are thus
\begin{gather}
  \boldsymbol{q} = q_1 \vect{e_z} + q_2 \vect{e_r} = - \gamma K(p_z) \left( \vect{e_z} p_z + \vect{e_r} p_r \right) \label{eq:darcy_velocity}\\
  \frac{\dd q_1}{\dd z} + \frac{1}{r} \frac{\dd}{\dd r} \left( r q_2 \right) = 0 \label{eq:continuity},
\end{gather}
where $\vect{e_z}$ and $\vect{e_r}$ describe the unit vectors in the $z$ and $r$ direction. The continuity of velocity is given in \eqref{eq:continuity}.

\begin{figure}
\centerline{\includegraphics{figures/pvs_model.pdf}}
\caption{Geometry of the perivascular drainage pathways model. The length $L$ of a typical artery in the brain can be assumed to be much greater than the width $W$ of the perivascular drainage pathways. Therefore we can approximate the flow as a lubrication model. The pulsation of the capillary is described by the function $h(z,t)$.\label{fig:pvs_model}}
\end{figure}

To determine the boundary conditions of the system consider the geometry in Figure \ref{fig:pvs_model}. Under the assumption that the vessel deformations are radially symmetric the positions of the inner and outer sheaths bounding the BM can be described by the two expressions $r = R_i(z,t)$ and $r = R_o(z,t)$, respectively. Assuming that the flows through the surfaces of the BM are negligible the following kinematic boundary condition applies
\beq \label{eq:kinematic}  
\frac{D}{D t} \left( r - R_i(z,t) \right) = 0 \quad \text{on } r = R_i(z,t)
\eeq
and analogously for $R_o(z,t)$.

Because the BM thickness is much smaller than the vessel length it is possible, on using the lubrication approximation, to derive a 1D model (in $z$) for the BM thickness $h_\textrm{bm}(z,t)$; this takes the form
\beq
\frac{\dd}{\dd t} \left( \gamma R_i(z,t) \cdot h_\textrm{bm}(z,t) \right) = \frac{\dd}{\dd z} \left( R_i(z,t) \cdot h_\textrm{bm}(z,t) \cdot K(p_{z}) \cdot p_{z} \right),\label{eq:pvs_model}
\eeq
where $\gamma$ is the fluid volume fraction in the BM. Equation \eqref{eq:pvs_model} describes the evolution of the BM thickness and can be solved using a finite difference approximation. For a detailed derivation see Supplementary Information 1. Another method to arrive at the same equation is derived by considering an infinitesimally small volume of BM $dV = 2\pi \cdot R_i(z,t) \cdot h_\textrm{bm}(z,t) \cdot dz$. Then the difference between the flux into and out of the element $dV$ is equal to $\dd (dV)/ \dd t$, which can be written as
\begin{align}
2\pi \frac{\dd}{\dd t} \left( \gamma R_i(z,t) \cdot h_\textrm{bm}(z,t) \right) &= \nonumber \\ 
2\pi \cdot \gamma R_i(z,t) \cdot h_\textrm{bm}(z,t) & \cdot q(z,t) - 2\pi \cdot \gamma R_i(z + dz,t) \cdot h_\textrm{bm}(z + dz,t) \cdot q(z + dz,t),
\end{align}
which is equal to the model equation \eqref{eq:pvs_model}. The pressure $p(z,t)$ and the displacement of the BM wall $R_i(z,t)$ are, to a very good approximation, determined solely by the arterial pressure (\ie almost independent of the BM flow), see Supplementary Information 2 for more details. $R_i(z,t)$ and $p_z$ can be obtained from a numerical artery model described in the next section. To model the protein valve that is being proposed here $K(p_z)$ is defined as a step function
\begin{equation}
  K(p_z) = \begin{cases}
    K_0 & p_z < 0\\
    K_1 & p_z \geq 0,
  \end{cases}
\end{equation}
chosen such that $K_1 > K_0 > 0$ so that the BM is more permeable to flow of ISF in the negative $z$-direction than in the positive one.

\section{Numerical Model of a Straight Arterial Section}

\begin{table}
\center
\caption{Geometry parameters of the MCA section modelled using VaMpy \cite{Diem2016a}.\label{tab:geometry}}
\begin{tabular}{ccc}
\toprule
Parameter & Value & Physical meaning\\
\midrule
\multicolumn{3}{l}{MCA}\\
\midrule
$R_u$ & \SI{0.14}{\centi\metre} & upstream radius\\
$R_d$ & \SI{0.14}{\centi\metre} & downstream radius\\
$L$ & \SI{7}{\centi\metre} & length\\
\midrule
\multicolumn{3}{l}{Left daughter vessel}\\
\midrule
$R_u$ & \SI{0.13}{\centi\metre} & upstream radius\\
$R_d$ & \SI{0.08}{\centi\metre} & downstream radius\\
$L$ & \SI{6.5}{\centi\metre} & length\\
\midrule
\multicolumn{3}{l}{Right daughter vessel}\\
\midrule
$R_u$ & \SI{0.13}{\centi\metre} & upstream radius\\
$R_d$ & \SI{0.08}{\centi\metre} & downstream radius\\
$L$ & \SI{6.4}{\centi\metre} & length\\
\bottomrule
\end{tabular}
\end{table}

In order to obtain the input functions $R_i(z,t)$ and $p_z$ for the analytical model developed in the previous section, a section of human MCA was modelled as a 1D axisymmetric tube in a cylindrical coordinate system with cross-sectional lumen area $A(z,t)$ and flux $q(z,t)$ using the Python package VaMpy \cite{Diem2016a}. The geometry parameters are listed in \autoref{tab:geometry}. Following the derivations in \cite{Olufsen2000,Diem2016} blood flow through arteries is represented by the conservation law
\begin{equation}
  \frac{\dd \boldsymbol{U}}{\dd t} + \frac{\dd \boldsymbol{F}}{\dd z} = \boldsymbol{S}, \label{eq:conservation}
\end{equation}
where
\begin{gather*}
  \boldsymbol{U} = \begin{pmatrix} A(z,t) \\ q(z,t) \end{pmatrix}, \quad \boldsymbol{F} = \begin{pmatrix} q(z,t)\\ \dfrac{q(z,t)^2}{A(z,t)} + f(r_0) \sqrt{A_0(z) A(z,t)} \end{pmatrix}, \quad \boldsymbol{S} = \begin{pmatrix} 0\\ S_1 \end{pmatrix}\\
  \begin{split}
    S_1 = -&\frac{2 \pi R(z,t)}{\delta_b \mathrm{Re}} \frac{q(z,t)}{A(z,t)} +\\
    &\left( 2 \sqrt{A(z,t)} \left( \sqrt{\pi} f(r_0) + \sqrt{A_0(z)} \frac{df(r_0)}{dr_0 } \right) - A(z,t) \frac{df(r_0)}{dr_0} \right) \frac{dr_0(z)}{dz}.
  \end{split}
\end{gather*}
The equation is solved for cross-sectional area $A(z,t)$ of the vessel lumen and flux $q(z,t)$ where $R(z,t) = q(z,t)/A(z,t)$ is the lumen radius, $\delta_b$ is the boundary layer thickness and $\text{Re}$ is the Reynolds number. Elasticity of the artery wall is modelled via the quantity $f(r_0) = 4Eh/(3r_0)$, where $E$ is the Young's modulus and $h$ is the wall thickness. The relationship $Eh/r_0$ is based on compliance estimates
\begin{equation}
  \frac{Eh}{r_0} = k_1 \exp \left( k_2 r_0(z) \right) + k_3
\end{equation}
with $k_1, k_2, k_3$ constants \cite{Olufsen2000}. Blood pressure $p(z,t)$ is related to $A(z,t)$ and $f(r_0)$ via the state equation
\beq
  p(z,t) - p_0 = f(r_0) \left( 1 - \frac{A_0(z)}{A(z,t)} \right),
\eeq
where $p_0$ is the diastolic pressure and $A_0(z)$ is the lumen radius at rest. Blood pressure provides the basis for determining ISF pressure inside the BM via the calculations in Supplementary Information 2.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This is just an example/guide for you to refer to when submitting manuscripts to Frontiers, it is not mandatory to use Frontiers .cls files nor frontiers.tex  %
% This will only generate the Manuscript, the final article will be typeset by Frontiers after acceptance.                                                 %
%                                                                                                                                                         %
% When submitting your files, remember to upload this *tex file, the pdf generated with it, the *bib file (if bibliography is not within the *tex) and all the figures.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Version 3.1 Generated 2015/22/05 %%%
%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
%%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%

%\documentclass{frontiersSCNS} % for Science, Engineering and Humanities and Social Sciences articles
%\documentclass{frontiersHLTH} % for Health articles
\documentclass{frontiersFPHY} % for Physics and Applied Mathematics and Statistics articles

%\setcitestyle{square}
\usepackage{url,hyperref,lineno,microtype}
\usepackage[onehalfspacing]{setspace}
\usepackage{textgreek}
\usepackage{xspace}
\usepackage{siunitx}
\linenumbers

\DeclareSIUnit\Molar{\textsc{M}}
\newcommand{\Ab}{A\textbeta\xspace}
\newcommand{\ie}{i.~e.\xspace}
\newcommand{\q}{\boldsymbol{q}}
\newcommand{\dd}{\partial}
\newcommand{\vect}[1]{\boldsymbol{#1}}
\newcommand{\Abs}{A\textbeta\textsubscript{1-40}\xspace}

% Leave a blank line between paragraphs instead of using \\

\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Diem {et~al.}} %use et al only if is more than 1 author
\def\Authors{Alexandra K. Diem\,$^{1,*}$, Neil W. Bressloff\,$^{1}$, Roxana O. Carare\,$^{2}$, Matthew MacGregor Sharp\,$^{2}$ and Giles Richardson\,$^3$}
% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
\def\Address{$^{1}$Computational Engineering and Design, Faculty of Engineering \& the Environment, University of Southampton, Southampton SO16 7QF, UK \\
$^{2}$Clinical Neurosciences, Faculty of Medicine, University of Southampton, Southampton General Hospital, Southampton SO16 6YD, UK\\
$^{3}$Mathematical Sciences, Faculty of Social, Human and Mathematical Sciences, University of Southampton, Southampton SO17 1BJ, UK}
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{A.~K.~Diem}
\def\corrAddress{Computational Engineering and Design, Faculty of Engineering \& the Environment, University of Southampton, Southampton SO16 7QF, UK}
\def\corrEmail{A.K.Diem@soton.ac.uk}

\begin{document}
\onecolumn
\firstpage{1}

\title[Arterial pulsations cannot drive periarterial drainage]{Arterial pulsations cannot drive intramural periarterial drainage: Significance for Alzheimer's disease}

\author[\firstAuthorLast ]{\Authors} %This field will be automatically populated
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated

\extraAuth{}% If there are more than 1 corresponding author, comment this line and uncomment the next one.
%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2, Organization X2, Street X2, City X2 , State XX2 (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% The sections below are for reference only.
%%%
%%% For Original Research Articles, Clinical Trial Articles, and Technology Reports the section headings should be those appropriate for your field and the research itself. It is recommended to organize your manuscript in the
%%% following sections or their equivalents for your field:
%%% Abstract, Introduction, Material and Methods, Results, and Discussion.
%%% Please note that the Material and Methods section can be placed in any of the following ways: before Results, before Discussion or after Discussion.
%%%
%%%For information about Clinical Trial Registration, please go to http://www.frontiersin.org/about/AuthorGuidelines#ClinicalTrialRegistration
%%%
%%% For Clinical Case Studies the following sections are mandatory: Abstract, Introduction, Background, Discussion, and Concluding Remarks.
%%%
%%% For all other article types there are no mandatory sections.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}

%%% Leave the Abstract empty if your article falls under any of the following categories: Editorial Book Review, Commentary, Field Grand Challenge, Opinion or specialty Grand Challenge.
  \section{
    Alzheimer's Disease (AD) is the most common form of dementia and to date there is no cure or efficient prophylaxis. The cognitive decline correlates with the accumulation of amyloid-\textbeta\xspace (\Ab) in the walls of capillaries and arteries. Our group has demonstrated that interstitial fluid and \Ab are eliminated from the brain along the basement membranes of capillaries and arteries, the intramural periarterial drainage (IPAD) pathway. With advancing age and arteriosclerosis, the stiffness of arterial walls, this pathway fails in its function and \Ab accumulates in the walls of arteries. In this study we tested the hypothesis that arterial pulsations drive IPAD and that a valve mechanism ensures the net drainage in a direction opposite to that of the blood flow. This hypothesis was tested using a mathematical model of the  drainage mechanism. We demonstrate firstly that arterial pulsations are not strong enough to produce drainage velocities comparable to experimental observations. Secondly, we demonstrate that a valve mechanism such as directional permeability of the IPAD pathway is necessary to achieve a net reverse flow. The mathematical simulation results are confirmed by assessing the pattern of IPAD in mice using pulse modulators, showing no significant alteration of IPAD. Our results indicate that forces other than the cardiac pulsations are responsible for efficient IPAD.
  }
%As a primary goal, the abstract should render the general significance and conceptual advance of the work clearly accessible to a broad readership. References should not be cited in the abstract.
%For full guidelines regarding your manuscript please refer to \href{http://www.frontiersin.org/about/AuthorGuidelines}{Author Guidelines} \\ or \textbf{Table \ref{Tab:01}} for a summary according to article type.

\tiny
 \keyFont{ \section{Keywords:} Alzheimer's disease; cerebral blood flow; perivascular drainage; intramural periarterial drainage; cerebral lymphatics } %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
\end{abstract}

\section{Introduction}

Alzheimer's disease (AD) is the commonest form of dementia. Although it has been studied for over 100 years \cite{Selkoe2001}, to date there is no cure as the processes relevant to its onset and progression have not yet been fully understood. The pathology of AD comprises the degeneration and death of neurons, synapse loss, neuroinflammation and the intracellular as well as extracellular accumulation of proteins \cite{Koffie2011}. The protein amyloid-\textbeta\xspace (\Ab) is produced as part of the normal metabolism of healthy brains, but it plays a major role in the development of AD \cite{Haass1992}. Its accumulation in the walls of cerebral blood vessels as cerebral amyloid angiopathy (CAA) is closely associated with the onset and progression of AD. CAA strongly correlates with the progression of AD and the degree of dementia as it appears in \SIrange{90}{96}{\percent} of AD patients \cite{Weller2006, Weller2009}.

In order to process metabolites, the brain requires a lymphatic system that transports waste products to the lymph nodes, where they are decomposed. While the rest of the body possesses a dedicated network of lymphatic vessels these do not exist in the brain \cite{Weller2010}. There is ample evidence that the basement membranes (BM) in the walls of cerebral arteries provide the clearance pathway for interstitial fluid (ISF) towards the surface of the brain, also referred to as intramural periarterial drainage (IPAD) \cite{Weller2010,Carare2008,Hawkes2011,Morris2014}. It is important not to confuse the IPAD pathway with the Virchow-Robin spaces, which are often referred to as perivascular (or ``paravascular'') spaces, possibly artefactual, located between the glia limitans and the arterial wall. Virchow-Robin spaces are virtual spaces and have no role in \Ab drainage. The BM forming the IPAD does not actually constitute a space, but rather a fluid-filled protein matrix. Our understanding of the mechanism that drives perivascular drainage through the BM is still very limited. The evidence indicates that ISF enters the BM at the capillary level and flows towards the lymph nodes in the neck via the BM cerebral arteries. This suggests that the counterintuitive phenomenon of ISF flow occurs in the reverse direction to blood flow \cite{Carare2008}.

A widely accepted hypothesis for the driving mechanism of this reverse perivascular drainage of solutes from the brain is that it is driven by arterial pulsations \cite{Weller2009,Carare2008,Hawkes2011,Morris2014,Schley2006,Attems2011,Wang2011,Iliff2012,Asgari2015,Sharp2015,Weller2015a}. In this paper we tested this hypothesis by using a multi-scale modelling approach combining analytical and numerical methods. The BM is modelled as a porous medium using a lubrication approximation. Flow inside the BM is driven by both the ISF pressure gradient and the deformation of the arterial wall, which are obtained from a numerical model of the middle cerebral artery (MCA). To give IPAD in the reverse direction to the blood flow, a valve mechanism is required; here it is modelled by a pressure gradient dependent permeability. It is shown that this valve mechanism is necessary to achieve net reverse drainage. However, due to the long wavelength of arterial pulsations the pressure gradients inside the BM are too small to drive fluid flow at the expected velocities. The resulting velocity of the model was at least four orders of magnitude smaller than values obtained experimentally \cite{Carare2008}. Therefore, arterial pulsations are concluded to be insufficient to drive perivascular drainage along the BM of cerebral arteries. The simulation results are confirmed by an experimental assessment of the effect of pulse modulation in a mouse model, which showed no significant difference in IPAD compare to control mice.

%\begin{methods}
\section{Numerical modelling of perivascular drainage of \Ab}

First, IPAD under a pressure-pulse driven scenario is evaluated using computational modelling. An analytical model of the BM was developed, which captures the preference of flow in the reverse direction to the blood flow. Pressure inside the BM is evaluated using a numerical model of blood pressure inside the artery and a stress analysis in the artery wall. These were then used to evaluate the velocity of perivascular drainage through the MCA.

\subsection{Porous medium model of lymphatic drainage through the BM}
\label{sec:diodemodel}

We propose a model of lymphatic drainage from the brain via the cerebral BM based on Darcy's law for flow in porous media. By exploiting the discrepancy of scales between the width of the BM and its length we approximate the flow by a one-dimensional lubrication model (see for example \cite{Ockendon1995}). Furthermore, we allow the permeability of the BM to depend on the pressure gradient of the ISF. Such a pressure gradient dependent permeability could, for instance, model proteins in the BM protein network that have structural properties that allow them to bend easily only in one direction, thereby acting as a valve, and thus provide a higher resistance to flow in the direction of the blood flow. A similar idea has been investigated in \cite{Sharp2015}, where proteins were explicitly modelled as cilia. However, here we model the active component of the BM generically in this model and thus allow for alternative explanations of the phenomenon.

The BM consists of a complex matrix of proteins (extracellular matrix, ECM) that can be interpreted as a porous medium through which soluble metabolites drain via the ISF. We model flow through this medium, in the standard fashion, by Darcy's law in a 2D cylindrical coordinate system. Darcy's law states that the flux $\q$ is related to the fluid pressure gradient $\boldsymbol{\nabla} p$ by
\begin{equation}\label{eq:darcy}
  \q = -\frac{k}{\mu} \boldsymbol{\nabla} p
\end{equation}
where $k$ is the intrinsic permeability of the ECM and $\mu$ is the viscosity of ISF. Since ISF flows up the artery, counter to the direction of arterial pulsations, a valve-like mechanism in the ECM is required if arterial pulsations are indeed the motive force for this flow. This valve-like property of the ECM can be modelled by a Darcy flow \eqref{eq:darcy}, in which the intrinsic permeability is a function of the axial pressure gradient $\dd p / \dd z$ along the vessel. On assuming an artery aligned along the $z$-axis with net blood flow and arterial pulsations propagating the positive $z$-direction we model valve-like properties of the ECM in the BM by a pressure gradient dependent permeability
\begin{equation}
k = \mu K(p_z) \qquad \text{where } p_z = \frac{\dd p}{\dd z}.
\end{equation}
In order to obtain a net flux up the vessel, counter to the blood flow, we require this function to be increasing in $p_z$, representing the notion that it is harder to push a flow in the positive $z$-direction ($p_z < 0$ and $K(p_z)$ small) than in the negative $z$-direction ($p_z \geq 0$ and $K(p_z)$ large).

The governing equations in cylindrical coordinates are thus
\begin{gather}
  \q = q_1 \vect{e_z} + q_2 \vect{e_r} = - K(p_z) \left( \vect{e_z} p_z + \vect{e_r} p_r \right) \label{eq:darcy_velocity}\\
  \frac{\dd q_1}{\dd z} + \frac{1}{r} \frac{\dd}{\dd r} \left( r q_2 \right) = 0 \label{eq:continuity},
\end{gather}
where $\vect{e_z}$ and $\vect{e_r}$ describe the unit vectors in the $z$ and $r$ direction. Incompressibility of the fluid is represented by \eqref{eq:continuity}.

To determine the boundary conditions of the system consider the geometry in Figure \ref{fig:pvs_model}. Under the assumption that the vessel deformations are radially symmetric the positions of the inner and outer sheaths bounding the BM can be described by the two expressions $r = R_i(z,t)$ and $r = R_o(z,t)$, respectively. Assuming that the flows through the surfaces of the BM are negligible the following kinematic boundary condition applies
\begin{equation} \label{eq:kinematic}  
\frac{D}{D t} \left( r - R_i(z,t) \right) = 0 \quad \text{on } r = R_i(z,t)
\end{equation}
and analogously for $R_o(z,t)$.

Because the BM thickness is much smaller than the vessel length it is possible, on using the lubrication approximation, to derive a 1D model (in $z$) for the BM thickness $h_\textrm{bm}(z,t)$; this takes the form
\begin{equation}
\begin{split}
\frac{\dd}{\dd t} \left( \gamma R_i(z,t) \cdot h_\textrm{bm}(z,t) \right) &=\\
\frac{\dd}{\dd z} \left( R_i(z,t) \right. & \left. \cdot h_\textrm{bm}(z,t) \cdot K(p_{z}) \cdot p_{z} \right),\label{eq:pvs_model}
\end{split}
\end{equation}
where $\gamma$ is the fluid volume fraction in the BM (see also SI~1). Equation \eqref{eq:pvs_model} describes the evolution of the BM thickness and can be solved using a finite difference approximation. For a detailed derivation see SI~1. Another method to arrive at the same equation is derived by considering an infinitesimally small volume of BM $dV = 2\pi \cdot R_i(z,t) \cdot h_\textrm{bm}(z,t) \cdot dz$. Then the difference between the flux into and out of the element $dV$ is equal to the rate of change of the fluid volume $\dd (\gamma dV)/ \dd t$, which can be written as
\begin{equation}
\begin{split}
\frac{\dd}{\dd t} \left( \gamma R_i(z,t) \cdot h_\textrm{bm}(z,t) \cdot dz \right) &=\\
& R_i(z,t) \cdot h_\textrm{bm}(z,t) \cdot q_1(z,t) -\\
R_i(z + dz,t) & \cdot h_\textrm{bm}(z + dz,t) \cdot q_1(z + dz,t),
\end{split}
\end{equation}
which is equal to the model equation \eqref{eq:pvs_model} in the limit $dz \rightarrow 0$. The pressure $p(z,t)$ and the displacement of the BM wall $R_i(z,t)$ are, to a very good approximation, determined solely by arterial pressure (\ie almost independent of the BM flow), see SI~2 for more details. $R_i(z,t)$ and $p_z$ are obtained from a numerical artery model described in the next section. To model the protein valve that is being proposed here, $K(p_z)$ is defined as a step function
\begin{equation}
  K(p_z) = \begin{cases}
    K_0 & p_z < 0\\
    K_1 & p_z \geq 0,
  \end{cases}
\end{equation}
such that $K_1 > K_0 > 0$ so that the BM is more permeable to ISF flow in the negative $z$-direction than in the positive one.

\subsection{Numerical Model of a Straight Arterial Section}

\begin{table}[!t]
\textbf{\refstepcounter{table}\label{tab:geometry} Table \arabic{table}.}{ Geometry parameters of the MCA section modelled using VaMpy \cite{Diem2016a}. }
\processtable{ }
             {\begin{tabular}{ccccl}\toprule
& MCA & left daughter & right daughter & \\
&     & vessel        & vessel         & \\
\midrule
$R_u$ & \SI{0.14}{\centi\metre} & \SI{0.13}{\centi\metre} & \SI{0.13}{\centi\metre} & upstream radius \\
$R_d$ & \SI{0.14}{\centi\metre} & \SI{0.08}{\centi\metre} & \SI{0.08}{\centi\metre} & downstream radius \\
$L$ & \SI{7}{\centi\metre} & \SI{6.5}{\centi\metre} & \SI{6.4}{\centi\metre} & length\\
\botrule
\end{tabular}}{}
\end{table}
